Viscous Erosion in Porous Media

### 2018-02-24 docs/methods/erosion_methods.tex
Adding some references

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Erosion is a fluid-mechanical process responsible for a variety
of phenomena in geophysical and other settings.  For example,
wind erosion causes sand dunes to be  longer on the windward
side~\cite{han1969}, while stream-bed erosion is responsible for branching structures
of rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is
also important in biophysical applications, for example plaque erosion in arteries
is key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}.  Another application,
which motivates the current work, is erosion caused by groundwater flow.
Here, a slow moving fluid erodes channels in the ground, and this
affects the fluid transport.  In all these examples, there is a feedback
process between the flow and the eroding bodies---the shape of the
bodies affect the flow, while the flow changes the shape by fluidizing
each body.  This fluid-mechanical feedback system has been simulated in
idealized cases such as a single eroding two-dimensional body in a
uniform flow~\cite{moo-ris-chi-zha-she2013} or a single mobile eroding
three-dimensional body~\cite{mit-spa2016}. However, to better understand
the effect of erosion on groundwater flow, multiply-connected domains
must be considered.  Here we simulate two-dimensional erosion, in
multiply-connected domain such as the porous geometry of many
groundwater flows (Fig.~\ref{fig:50bodies}).

@@

The main idea is to introduce an artificial tangential velocity that
keeps collocation points equally spaced with respect to arc length;
i.e.~the grid of $\alpha$ should always remain equally spaced. The
tangential velocity that does the trick satisfies~\cite{hou-low-she1994}
\begin{align*}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}.
\end{align*}
The interface evolution is then given by
\begin{align*}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{align*}
This evolution equation can be recast in the variables $\theta$ and $L$
as
\begin{align*}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align*}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
System \eqref{eqn:Levo1}--\eqref{eqn:thetaevo1} comprise the main evolution equations that we will solve numerically in Section~\ref{sec:timeStepping}. In addition, it is necessary to track a reference point for each body, so that the variables $\theta$ and $L$ can be converted back into a cartesian grid. We choose to track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
Tracking the surface mean is somewhat simpler than tracking the center of mass, but either choice would be valid.

@@

At the second stage of RK2, we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \, \zeta^{(n+1/2)} \right) \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we insert \eqref{eq:mu1d} into \eqref{eq:th1d} and solve for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Again, both exponential terms act to smooth high-frequency oscillations in shape. We note that the above approach would generalize to higher-order Runge-Kutta methods, such as RK4.

@@

A brief look at Fig.~\ref{shrink_intface}(a) reveals that
the spacing between successive interfaces becomes larger as the body
shrinks, suggesting that the overall shear stress is growing. Of course,
such a relationship is consistent with the well-known 3D Stokes drag
law, $\tau \sim \mu \umax / L$, since stress is inversely proportional
to body scale $L$. This law must be modified in two dimensions, though.
The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory. 
\todo[inline]{Here: add slender citations}
For a slender body in flow perpendicular to its long-axis $W \gg L$, the drag is proportional to
$\mu \umax W/ \log(W/L)$. This relationship implies that, for a 2D
cross-section, the stress per unit span can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{L \log(W/L)}.
\end{align}
In interpreting our 2D simulations, we will treat the span, $W$, as fixed with the
requirement $W \gg L$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}

Erosion is a fluid-mechanical process responsible for a variety
of phenomena in geophysical and other settings.  For example,
wind erosion causes sand dunes to be  longer on the windward
side~\cite{han1969}, while stream-bed erosion is responsible for branching structures
of rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is
also important in biophysical applications, for example plaque erosion in arteries
is key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}. 
Much recent work has been devoted to understanding fundamental aspects of the fluid-structure interaction that occurs during the erosion \cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, mit-spa2016, hewett2017evolution, moore2017riemann, lachaussee2018competitive, lopez2018cfd}, and related processes like dissolution or melting \cite{Huang2015, kondratiuk2015steadily, rycroft2016asymmetric, cohen2016erosion, yuan2017dynamics, claudin2017dissolution, hewett2017pear}. These processes can be linked conceptually as `sculpting with flow' \cite{ristroph2018sculpting}.

Another application,
which motivates the current work, is erosion caused by groundwater flow.
Here, a slow moving fluid erodes channels in the ground, and this
affects the fluid transport.  In all these examples, there is a feedback
process between the flow and the eroding bodies---the shape of the
bodies affect the flow, while the flow changes the shape by fluidizing
each body.  This fluid-mechanical feedback system has been simulated in
idealized cases such as a single eroding two-dimensional body in a
uniform flow~\cite{moo-ris-chi-zha-she2013} or a single mobile eroding
three-dimensional body~\cite{mit-spa2016}. However, to better understand
the effect of erosion on groundwater flow, multiply-connected domains
must be considered.  Here we simulate two-dimensional erosion, in
multiply-connected domain such as the porous geometry of many
groundwater flows (Fig.~\ref{fig:50bodies}).

@@

The main idea is to introduce an artificial tangential velocity that
keeps collocation points equally spaced with respect to arc length;
i.e.~the grid of $\alpha$ should always remain equally spaced. The
tangential velocity that does the trick satisfies~\cite{hou-low-she1994}
\begin{align*}
\pderiv{\Vs}{\alpha} = \thalpha \Vn - \mean{\thalpha \Vn}.
\end{align*}
The interface evolution is then given by
\begin{align*}
\dot{\xx}(t) = \Vn \nn + \Vs \ss.
\end{align*}
This evolution equation can be recast in the variables $\theta$ and $L$
as
\begin{align*}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align*}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
System \eqref{eqn:Levo1}--\eqref{eqn:thetaevo1} comprise the main evolution equations that we will solve numerically in Section~\ref{sec:timeStepping}.

In addition, it is necessary to track a reference point for each body, so that the variables $\theta$ and $L$ can be converted back into a cartesian grid. We choose to track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\label{eq:xsurf}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
Tracking the surface mean is somewhat simpler than tracking the center of mass, but either choice would be valid.

@@

At the second stage of RK2, we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \, \zeta^{(n+1/2)} \right) \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we insert \eqref{eq:mu1d} into \eqref{eq:th1d} and solve for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Again, both exponential terms act to smooth high-frequency oscillations in shape. We note that the above approach would generalize to higher-order Runge-Kutta methods, such as RK4. Finally, we track the reference point $\mean{x}$ of each body by simply discretizing \eqref{eq:xsurf} with the same RK2 method.

@@

Figure~\ref{shrink_intface}(a) shows that the spacing between successive interfaces becomes larger as the body shrinks, suggesting that the overall shear stress is growing. This observation is consistent with the well-known 3D Stokes drag law, $\tau \sim \mu \umax / L$, since stress is inversely proportional to body scale $L$. This law must be modified in two dimensions, though.
The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory \cite{poz1997, bouzarth2007epicyclic, camassa2008trajectory, MooreJFM2012}. 
\todo[inline]{Here: add slender citations}
For a slender body in flow perpendicular to its long-axis $W \gg L$, the drag is proportional to
$\mu \umax W/ \log(W/L)$. This relationship implies that, for a 2D
cross-section, the stress per unit span can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{L \log(W/L)}.
\end{align}
In interpreting our 2D simulations, we will treat the span, $W$, as fixed with the
requirement $W \gg L$.
